---
title: "fftab"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(digits = 3)
library(ggfortify)
library(patchwork)
library(ggplot2)
```

The `fftab` package provides a tidy interface for Fourier Transform operations by storing Fourier coefficients and their associated frequencies in a tabular format (`tibble`). This design facilitates the manipulation of spectral data using pipes and functions from the tidyverse.

We’ll demonstrate the package’s capabilities using reproducible examples.

### Setup

Load the package and set the random seed:

```{r setup, message=FALSE, error=FALSE, warning=FALSE}
library(fftab)
set.seed(1234)
```

The `fftab` function supports various input types, including numeric and complex vectors, time series objects, and multidimensional arrays.

---

### Vector Input

A simple example is the FFT of a numeric vector:

```{r}
v <- rnorm(8)
v_fft <- fftab(v)
print(v_fft)
```

The output `tibble` includes:

* `.dim_1`: Normalized frequencies in cycles per unit length. For an even-length vector, the maximum frequency is the Nyquist frequency (0.5). Frequencies wrap around zero, producing negative frequencies.
* `fx`: Fourier coefficients in complex form.

To facilitate manipulation, you can switch between representations—complex (`cplx`), rectangular (`rect`), and polar (`polr`):

```{r}
v_fft |> to_rect() |> print(n = 3)
v_fft |> to_polr() |> print(n = 3)
v_fft |> set_repr(c("polr", "rect", "cplx")) |> print(n = 3)
```

Retrieve individual components, such as phase or modulus, with `get_*` functions:

```{r}
v_fft |> get_arg()
```

These functions work seamlessly across all representations, ensuring consistent outputs.

**Warning:** When modifying an `fftab` object:

* If multiple representations (e.g., complex, polar, rectangular) are present, ensure consistency to avoid ambiguity during the inverse transform.
* Changes to specific columns, such as `arg`, do not automatically propagate to other representations.
* Adding or deleting rows can lead to nonsensical results when inverting the transform.

Always verify the structure and contents of an `fftab` object after modifications to ensure it remains valid.

---

### Complex Input

The `fftab` object tracks whether the input is real or complex, ensuring correct handling in inverse transforms:

```{r}
cplx <- complex(modulus = rnorm(4), argument = runif(4, 0, pi))
cbind(orig = cplx, reconst = cplx |> fftab() |> ifftab())
```

For real-valued input returns real-valued output when the transform is inverted.

```{r}
real_v <- runif(4)
cbind(orig = real_v, reconst = real_v |> fftab() |> ifftab())
```

---

### Time Series Input

Time series objects in R include implicit units and a frequency attribute. For example, the `sunspot.month` dataset represents monthly counts spanning from 1749 to 2014. The sampling frequency (12) corresponds to cycles per year:

```{r}
sunspot.month |> fftab() |> print(n = 5)
```

Here, `.dim_1` contains frequencies in cycles per year. It is important to understand the difference,
as the Nyquist frequency will be 6 cycles per year and not $1/2$ cycle per step as would be the case when analyzing a vector without frequency information.

The original data can be restored using `ifftab`:

```{r}
sunspot.month |> fftab() |> ifftab() |> str()
```

---

### Array Input

Arrays retain their dimensions through the Fourier Transform:

```{r}
ra <- matrix(rnorm(9), 3)
ra |> fftab() |> to_polr()
ra |> fftab() |> ifftab()
```

---

### Example 1: Applying a Phase Shift

The polar representation allows manipulation of phase information. Let’s create a sinusoid and examine it:

```{r}
#| fig-asp: 0.3333
s1 <- seq(-pi, pi, length.out = 200) |>
  tibble::as_tibble() |>
  dplyr::mutate(x = value, y = cos(4 * value), treatment = "original", .keep = "none")

ggplot(s1) +
  geom_line(aes(x = x, y = y)) +
  theme_classic()
```

Perform the FFT and plot the results:

```{r}
#| fig-asp: 0.3333
s1_fft <- with(s1, y |> fftab() |> to_polr())
plot(s1_fft) # Quick visualization
```

To shift the signal, adjust the phase (`arg`) while respecting conjugate symmetry:

```{r}
s2_fft <- s1_fft |>
  dplyr::mutate(arg = dplyr::case_when(
    .dim_1 == 0 ~ arg,      # DC component
    .dim_1 == 0.5 ~ arg,    # Nyquist frequency
    .dim_1 > 0 ~ arg - pi,  # Positive frequencies
    .dim_1 < 0 ~ arg + pi   # Conjugate negative frequencies
  ))
```

Reconstruct and compare the shifted signal to the original:

```{r}
#| fig-asp: 0.3333
s1 |>
  dplyr::mutate(y = s2_fft |> ifftab(), treatment = "shifted") |>
  dplyr::bind_rows(s1) |>
  ggplot() +
  geom_line(aes(x = x, y = y, color = treatment), lwd = 1, alpha = 0.5) +
  scale_color_manual(values = c("darkblue", "darkred")) +
  theme_classic()
```

---

### Example 2: Time Series Filtering

To denoise the `sunspot.month` data, we suppress high-frequency components:

```{r}
ggplot(fortify(sunspot.month)) +
  geom_line(aes(x = Index, y = Data)) +
  ylab("Sunspot count") +
  xlab("Year") +
  theme_classic() ->
p1

smoothed <- sunspot.month |>
  fftab() |>
  to_polr() |>
  dplyr::mutate(mod = ifelse(abs(.dim_1) > 0.2, 0, mod)) |>
  ifftab()

ggplot(fortify(smoothed)) +
  geom_line(aes(x = Index, y = Data)) +
  theme_classic() ->
p2

p1 / p2
```

The `dplyr::mutate` call zeros out the magnitudes of any components with
periodicity of 5 years or fewer.

---

### Example 3: 2D Gaussian Autocorrelation

A gaussian kernel in the frequency domain has mangnitudes
$$
\left | F \left ( \vec\omega \right) \right | = -\pi^2\sigma^2\left|\left| \vec\omega \right|\right| ^2
$$
where $\left|\left| \vec\omega \right|\right| ^2$ is the sum of squared moduli across the dimensions.
`fftab` provides functions for computing the L2 norm and squared L2 norm of the frequencies in any
number of dimensions allowing easy access to these numbers in computations.

```{r}
sigma <- 64
scale_fac <- -pi^2 * sigma^2

matrix(rnorm(512 * 512), 512) |>
  fftab() |>
  to_polr() |>
  add_l2sq() |>
  dplyr::mutate(mod = mod * exp(scale_fac * l2sq)) |>
  ifftab() ->
gauss_acf

tidyr::expand_grid(x = 1:512, y = 1:512) |>
  tibble::add_column(z = gauss_acf |> as.vector() |> scale()) |>
  ggplot() +
  aes(x = x, y = y, z = z) +
  geom_contour_filled(bins = 16) +
  geom_contour(bins = 16, color = "darkgrey") +
  coord_equal() +
  theme_void()
```

The result of this operation is a circular convolution of the white-noise input
with a gaussian kernel.

---

### Conclusion

The `fftab` package simplifies Fourier Transform operations by combining spectral coefficients and frequencies in a `tibble`. This integration enables seamless manipulation of frequency-domain data with tidyverse tools, making `fftab` a powerful resource for spectral analysis.

---
